Model Recognition as Planning

### 2018-09-21 doc/icaps-19/model-recognition.tex
paper icaps19: pages [1-5] stable. Ready for review

The intuition behind this compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the action model $\mathcal{M}$ to build $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that modifies the preconditions and effects of the action schemes in $\mathcal{M}$ using to the two {\em edit operations} defined above, {\em deletion} and {\em insertion}. 
\item {\bf Validates the edited model $\mathcal{M}'$}. The solution plan continues with a postfix that:
\begin{enumerate}
\item Induces an optimal solution plan $\pi^*_\tau$ for the original classical planning problem $P$.
\item Validates that $\tau$ is an observation of the execution of $\pi^*_\tau$ on the classical planning problem $P$.
\end{enumerate}
\end{enumerate}

Figure~\ref{fig:plan-pdistance} shows the plan with a prefix (steps [0,6]) for editing the planning model of Figure\ref{fig:model-example} when its action {\tt\small inc-x} is defined without preconditions and its positive/negative effects are swapped wrt Figure~\ref{fig:model-example}. The postfix of the plan (steps [7,13]) validates the edited action model at the observation of the plan execution illustrated at Figure~\ref{fig:grid-example}. 
\begin{figure}
{\tt\scriptsize
00 : (insert\_pre\_xcoord\_v1\_inc-x)\\
01 : (insert\_pre\_next\_v1\_v2\_inc-x)\\
02 : (insert\_pre\_q0\_inc-x)\\
03 : (delete\_del\_xcoord\_v2\_inc-x)\\
04 : (delete\_add\_xcoord\_v1\_inc-x)\\
05 : (insert\_del\_xcoord\_v1\_inc-x)\\
06 : (insert\_add\_xcoord\_v2\_inc-x)\\
07 : (editable\_inc-x 1 2)\\
08 : (editable\_inc-x 2 3)\\
09 : (editable\_inc-x 3 4)\\
10 : (editable\_inc-x 4 5)\\
11 : (editable\_inc-y-even 1 2)\\
12 : (editable\_dec-x 5 4)\\
13 : (validate\_1)
}
 \caption{\small Plan for editing (steps [0-6]) and validating (steps [7-13]) the planning model of Figure\ref{fig:model-example} when action {\tt\small inc-x} has no preconditions and positive/negative effects are swapped wrt Figure\ref{fig:model-example}.}
\label{fig:plan-pdistance}
\end{figure}

@@

The interest of having a propositional encoding for \strips\ action schema is that it allows to define {\em editable actions} that is, actions whose semantics is given by the value of these particular fluents on the current state. Formally given an operator schema $\xi\in\mathcal{M}$ its {\em editable} version can be compactly specified with {\em conditional effects}:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{editable_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\\
\cond(\mathsf{editable_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}.
\end{align*}
\end{small}
Since operators headers are given as input, the variables $pars(\xi)$ are bound to the objects in $\omega\subseteq\Omega^{ar(\xi)}$ that appear at the same position. Figure~\ref{fig:editable} shows the PDDL encoding of the {\em editable version} of the {\tt\small inc-x(?v1,?v2)} schema for robot navigation in a $n\times n$ grid (see Figure~\ref{fig:model-example}). Note that this editable schema, when the fluents of Figure~\ref{fig:encoding} hold, behaves exactly as defined in Figure~\ref{fig:model-example}. Further, it allows to swap the semantics of any two schema provided that they are {\em comparable}.

@@

\subsection{The compilation formalization}
Conditional effects allow us to compactly define our compilation. Given a \strips\ model $\mathcal{M}\in M$ and the observation $\tau$ of the execution of a plan for solving $P=\tup{F,A,I,G}$, our compilation outputs a classical planning task with conditionl effects $P'=\tup{F',A',I',G'}$ such that:
\begin{itemize}
\item $F'$ contains:
\begin{itemize}
\item The original fluents $F$. 
\item Fluents $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ modeling the space of \strips\ models. 
\item The fluents $F_{\pi}=\{plan(name(a_i),\Omega^{ar(a_i)},i)\}_{\small 1\leq i\leq n}$ to code the $i^{th}$ action in $\tau$. The static facts $next_{i,i+1}$ and the fluents $at_i$, {\small $1\leq i< n$}, are also added to iterate through the $n$ steps of $\tau$.
\item The fluents $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\tau$ where the action model is validated.
\item The fluents $mode_{edit}$ and $mode_{val}$ to indicate whether the operator schemas are edited or validated.
\end{itemize}
\item $I'$ extends the original initial state $I$ with the fluent $mode_{edit}$ set to true as well as the fluents $F_{\pi}$ plus fluents $at_1$ and $\{next_{i,i+1}\}$, {\small $1\leq i<n$}, for tracking the plan step where the action model is validated. Our compilation assumes that initially $\mathcal{M}'$ is defined as $\mathcal{M}$. Therefore fluents $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ hold as given by $\mathcal{M}$.

\item $G'=G\bigcup\{at_n,test_m\}$.
\item $A'$ comprises three kinds of actions with conditional effects:
\begin{enumerate}
\item Actions for {\em editing} operator schema $\xi\in\mathcal{M}$:
\begin{itemize}
\item Actions for adding a {\em precondition} $f\in F_v(\xi)$ from the action schema $\xi\in\mathcal{M}$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programPre_{f,\xi}})=&\{\neg pre_{f}(\xi), \neg del_{f}(\xi),\neg add_{f}(\xi), mode_{edit}\},\\
\cond(\mathsf{programPre_{f,\xi}})=&\{\emptyset\}\rhd\{pre_{f}(\xi)\}.
\end{align*}
\end{small}

\item Actions for adding a {\em negative} or {\em positive} effect $f\in F_v(\xi)$ to the action schema $\xi\in\mathcal{M}$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programEff_{f,\xi}})=&\{\neg del_{f}(\xi),\neg add_{f}(\xi), mode_{edit}\},\\
\cond(\mathsf{programEff_{f,\xi}})=&\{pre_{f}(\xi)\}\rhd\{del_{f}(\xi)\},\\
&\{\neg pre_{f}(\xi)\}\rhd\{add_{f}(\xi)\}.
\end{align*}
\end{small}
\end{itemize}
Besides these actions, the $A'$ set also contains the actions for {\em deleting} a precondition and a negative/positive effect.

\item Actions for {\em applying} an edited operator schema $\xi\in\mathcal{M}$ bound with objects $\omega\subseteq\Omega^{ar(\xi)}$. Since operators headers are given as input, the variables $pars(\xi)$ are bound to the objects in $\omega$ that appear at the same position. Figure~\ref{fig:editable} shows the PDDL encoding of the action for applying a programmed operator.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\\
&\cup \{\neg mode_{val}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{edit}\}\rhd\{\neg mode_{edit}\},\\
&\{\emptyset\}\rhd\{mode_{val}\}.
\end{align*}
\end{small}

When the observation $\tau$ includes observed actions, then the extra conditional effects $\{at_{i},plan(name(a_i),\Omega^{ar(a_i)},i)\}\rhd\{\neg at_{i},at_{i+1}\}_{\forall i\in [1,n]}$ are included in the $\mathsf{apply_{\xi,\omega}}$ actions to validate that actions are applied, exclusively, in the same order as they appear in $\tau$.\\

>>>>

The compilation is an extension of the classical planning compilation for the learning of \strips\ planning models~\cite{aineto2018learning}. The intuition behind this compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the action model $\mathcal{M}$ to build $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that modifies the preconditions and effects of the action schemes in $\mathcal{M}$ using to the two {\em edit operations} defined above, {\em deletion} and {\em insertion}. 
\item {\bf Validates the edited model $\mathcal{M}'$}. The solution plan continues with a postfix that:
\begin{enumerate}
\item Induces an optimal solution plan $\pi^*_\tau$ for the original classical planning problem $P$.
\item Validates that $\tau$ is an observation of the execution of $\pi^*_\tau$ on the classical planning problem $P$.
\end{enumerate}
\end{enumerate}

Figure~\ref{fig:plan-pdistance} shows the plan with a prefix (steps [0,6]) for editing the planning model of Figure\ref{fig:model-example} when its action {\tt\small inc-x} is defined without preconditions and its positive/negative effects are swapped wrt Figure~\ref{fig:model-example}. The postfix of the plan (steps [7,13]) validates the edited action model at the observation of the plan execution illustrated at Figure~\ref{fig:grid-example}. 
\begin{figure}
{\tt\footnotesize
00 : (insert\_pre\_xcoord\_v1\_inc-x)\\
01 : (insert\_pre\_next\_v1\_v2\_inc-x)\\
02 : (insert\_pre\_q0\_inc-x)\\
03 : (delete\_del\_xcoord\_v2\_inc-x)\\
04 : (delete\_add\_xcoord\_v1\_inc-x)\\
05 : (insert\_del\_xcoord\_v1\_inc-x)\\
06 : (insert\_add\_xcoord\_v2\_inc-x)\\
07 : (editable\_inc-x 1 2)\\
08 : (editable\_inc-x 2 3)\\
09 : (editable\_inc-x 3 4)\\
10 : (editable\_inc-x 4 5)\\
11 : (editable\_inc-y-even 1 2)\\
12 : (editable\_dec-x 5 4)\\
13 : (validate\_1)
}
 \caption{\small Plan for editing (steps [0-6]) and validating (steps [7-13]) the planning model of Figure\ref{fig:model-example} when action {\tt\small inc-x} has no preconditions and positive/negative effects are swapped wrt Figure\ref{fig:model-example}.}
\label{fig:plan-pdistance}
\end{figure}

@@

The interest of having a propositional encoding for \strips\ action schema is that it allows to define {\em editable actions} that is, actions whose semantics is given by the value of these particular fluents on the current state. Formally given an operator schema $\xi\in\mathcal{M}$ its {\em editable} version can be compactly specified with {\em conditional effects}:
\begin{small}  
\begin{align*}
\hspace*{7pt}\pre(\mathsf{editable_{\xi,\omega}})=&\{pre\_e\_name(\xi)\implies e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}\\
\cond(\mathsf{editable_{\xi,\omega}})=&\{del\_e\_name(\xi)\}\rhd\{\neg e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}},\\
&\{add\_e\_name(\xi)\}\}\rhd\{e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}.
\end{align*}
\end{small}
Since operators headers are given as input, the variables $pars(\xi)$ are bound to the objects that appear at the same position. Figure~\ref{fig:editable} shows the PDDL encoding of the {\em editable version} of the {\tt\small inc-x(?v1,?v2)} schema for robot navigation in a $n\times n$ grid (see Figure~\ref{fig:model-example}). Note that this editable schema, when the fluents of Figure~\ref{fig:encoding} hold, behaves exactly as defined in Figure~\ref{fig:model-example}. Further, it allows to swap the semantics of any two schema provided that they are {\em comparable}.

@@

\subsection{The compilation formalization}
Conditional effects allow us to compactly define our compilation. Given a \strips\ model $\mathcal{M}\in M$ and the observation $\tau$ of the execution of a plan for solving $P=\tup{F,A,I,G}$, our compilation outputs a classical planning task with conditionl effects $P'=\tup{F',A',I',G'}$ such that:
\begin{itemize}
\item $F'$ contains:
\begin{itemize}
\item The original fluents $F$. 
\item Fluents $[pre|del|add]\_e\_name(\xi)$ to model the possible \strips\ schema. 
\item The fluents to code the {\em observation constraints} and that are of these kinds:
\begin{itemize}
\item $F_{\pi}=\{plan(name(a_i),\Omega^{ar(a_i)},i)\}_{\small 1\leq i\leq n}$ to code the $i^{th}$ action in $\tau$. The static facts $next_{i,i+1}$ and the fluents $at_i$, {\small $1\leq i< n$}, are also added to iterate through the $n$ steps of $\tau$.
\item The fluents $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\tau$ where the action model is validated.
\end{itemize}
\item The fluents $mode_{edit}$ and $mode_{val}$ to indicate whether the operator schemas are edited or validated.
\end{itemize}
\item $I'$ extends the original initial state $I$ with the fluent $mode_{edit}$ set to true as well as the fluents $F_{\pi}$ plus fluents $at_1$ and $\{next_{i,i+1}\}$, {\small $1\leq i<n$}, for tracking the plan step where the action model is validated. Our compilation assumes that initially $\mathcal{M}'$ is defined as $\mathcal{M}$. Therefore fluents $[pre|del|add]\_e\_name(\xi)$ hold as given by $\mathcal{M}$.

\item $G'=G\bigcup\{at_n,test_m\}$.
\item $A'$ comprises three kinds of actions with conditional effects:
\begin{enumerate}
\item The {\em editable} version of the original actions in $A$. This actions have now  an extra preconditions  becuase they can only be applied in the {\em edition} mode (i.e. when $mode_{edit}$ holds). When the observation $\tau$ includes observed actions, they also include the extra conditional effects $\{at_{i},plan(name(a_i),\Omega^{ar(a_i)},i)\}\rhd\{\neg at_{i},at_{i+1}\}_{\forall i\in [1,n]}$ to validate that actions are applied, exclusively, in the same order as they appear in $\tau$.\\

\item Actions for {\em editing} operator schema $\xi\in\mathcal{M}$:
\begin{itemize}
\item Actions for inserting a into an action schema $\xi\in\mathcal{M}$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programPre_{e,\xi}})=&\{\neg pre\_e\_name(\xi), \neg del\_e\_name(\xi),\\
& \neg add\_e\_name(\xi), mode_{edit}\},\\
\cond(\mathsf{programPre_{e,\xi}})=&\{\emptyset\}\rhd\{pre\_e\_name(\xi)\}.
\end{align*}
\end{small}

\item Actions for inserting adding a {\em negative} or {\em positive} effect into the action schema $\xi\in\mathcal{M}$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programEff_{e,\xi}})=&\{\neg del\_e\_name(\xi),\\
& \neg add\_e\_name(\xi),mode_{edit}\},\\
\cond(\mathsf{programEff_{e,\xi}})=&\{pre\_e\_name(\xi)\}\rhd\{del\_e\_name(\xi)\},\\
&\{\neg pre\_e\_name(\xi)\}\rhd\{add\_e\_name(\xi)\}.
\end{align*}
\end{small}
\end{itemize}
Besides these actions, the $A'$ set also contains the actions for {\em deleting} a precondition and a negative/positive effect.
